---
title: "Disease pyramid"
subtitle: "Preparing the data"
author: "Matan Markfeld"
date: "Last edit: 2026-08-24"
output: 
  html_document:
    toc: true
    toc_float: true
    collapse: false
    number_sections: true
    code_folding: hide
    highlight: tango
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "asis", message=FALSE, warning=FALSE, cache=TRUE, eval = TRUE, dev = c('png'), out.width = '100%', out.height='100%')
```


```{r load libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
library(ggplot2)
library(tidytext)
library(vegan)
library(reshape2)
library(GUniFrac)
library(RColorBrewer)
library(picante)

rm(list=ls())
```

# Overview

This document assembles the host-level predictor table used in the main analysis
from raw field, environmental, and microbiome data. For each host individual it
prepares three groups of variables:

* **Host traits** — body mass, structural body size, body condition (BCI), and sex.
* **Environmental / land-use descriptors** — a habitat-structure PCA that combines
  vegetation structure, small-mammal density, and distance to the village centre.
* **Microbiome measures** — Shannon diversity, Faith's phylogenetic diversity,
  weighted-UniFrac uniqueness, and the Firmicutes/Bacteroidota log-ratio.

All continuous variables are z-scored *within each host species* so the two focal
species sit on a common scale. The set of analysis hosts is defined by the hosts
present in the parasite table.

**Input files** (under `../data/`):

* `data_raw/small_mammals/Terrestrial_Mammals.csv`, `Trap_Plots.csv`, `Trap_Grids.csv`
* `data_processed/parasites_long_10_lulu.csv`
* `data_processed/data_microbiome_rra0.001_th5000_2.csv`
* `data_processed/ASV_tree_ML_rooted.rds`
* `data_raw/microbiome/ASVs_taxonomy_new.tsv`

**Output files** (under `../data/data_processed/`):

* `data_x_final.csv` — one row per host with all predictors and site metadata.
* `data_rand_final.csv` — host and site grouping variables (for random effects).

# Data

```{r}
# reading the data
host_sp <- c("Rattus rattus", "Microgale brevicaudata")

data_sm <- read_csv("../data/data_raw/small_mammals/Terrestrial_Mammals.csv") %>% 
  filter(grepl("^TMR", animal_id)) %>% 
  mutate(host_ID = as.numeric(gsub(".*?([0-9]+).*", "\\1", animal_id))) %>% 
  dplyr::select(host_ID, field_identification, village, habitat_type, season, dist_village_center, sex, mass, hb, hf) %>%
  dplyr::rename(host_species = field_identification, grid = habitat_type) %>% 
  mutate(grid = factor(grid, levels = c("semi-intact_forest","secondary_forest","brushy_regrowth","agriculture","flooded_rice","agroforest","village"))) %>%
  filter(host_species %in% host_sp)

data_parasites <- read_csv("../data/data_processed/parasites_long_10_lulu.csv")
data_microbiome <- read_csv("../data/data_processed/data_microbiome_rra0.001_th5000_2.csv")
```

```{r}
# host IDs we use in the analysis
host_ids <- unique(data_parasites$host_ID)
```

# Host traits
Four host attributes are prepared (all continuous ones are z-scored *within each species*
so the two very different focal species sit on a common scale):

* **mass_z**  – body mass [g], z-scored per species (kept for flexibility. overlaps with size + BCI).
* **size_z**  – structural body size = log(hindfoot length), z-scored per species.
                Hindfoot is a more repeatable skeletal ruler than head-body length.
* **BCI**     – body condition = OLS residual of log(mass) on log(hindfoot), fit per species,
                then z-scored per species. Orthogonal to size by construction.
* **sex**     – 1 = male, 0 = female.

```{r}
host_traits <- data_sm %>% 
  filter(host_ID %in% host_ids) %>% 
  select(host_ID, host_species, village, grid, season, mass, sex, hb, hf) %>% 
  mutate(sex = ifelse(sex=="male",1,0)) %>% 
  mutate(season = as.character(season)) %>% 
  mutate(log_mass = log(mass), log_hb = log(hb), log_hf = log(hf)) %>% 
  group_by(host_species) %>%
  mutate(mass_z = as.numeric(scale(mass)),        # body mass, z-scored per species
         size_z = as.numeric(scale(log_hf))) %>%  # structural size (hindfoot), z-scored per species
  ungroup()
```

## BCI
```{r}
# Body condition index (BCI):
# OLS residual of log(mass) on log(hindfoot), fit within each species. OLS residuals are
# orthogonal to the size predictor by construction, so BCI does not correlate with size_z
# Residuals are then z-scored per species for cross-species comparability.
calculate_bci <- function(df) {
  model <- lm(log_mass ~ log_hf, data = df, na.action = na.exclude)
  df$BCI_resid <- resid(model)
  return(df)
}

# Apply function separately by host species, then z-score the residual per species
host_bci <- host_traits %>%
  group_by(host_species) %>%
  group_modify(~ calculate_bci(.x)) %>%
  group_by(host_species) %>%
  mutate(BCI = as.numeric(scale(BCI_resid))) %>%
  ungroup() %>% 
  select(host_ID, BCI)

# plotting
#pdf('../results/BCI.pdf')
host_traits %>% 
  drop_na(log_mass, log_hf) %>% 
  ggplot(aes(x = log_hf, y = log_mass)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth=1) +
  facet_wrap(~ host_species) +
  labs(
    x = "log(Hindfoot) [mm]",
    y = "log(Body Mass) [g]"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold")
  )
#dev.off()
```

## Final table
```{r}
x1_host_traits <- host_traits %>% 
  select(host_ID, host_species, village, grid, season, mass, mass_z, size_z, sex) %>% 
  left_join(host_bci, by="host_ID")
```


# Environmental variables

## Grid vegetation processing

PCA of vegetation per village per grid per season

```{r}
habitat.raw <- read_csv("../data/data_raw/small_mammals/Trap_Plots.csv") %>% 
  mutate(season = ifelse(is.na(season), 3, season)) %>% 
  filter(village == "Mandena")

habitat <- habitat.raw %>% 
  mutate(grid_name = case_when(grid_name=="Primary forest" ~ "semi-intact_forest",
                               grid_name=="Degraded forest" ~ "semi-intact_forest",
                               grid_name=="Secondary forest" ~ "secondary_forest",
                               grid_name=="Savoka" ~ "brushy_regrowth",
                               grid_name=="Sugar cane" ~ "agriculture",
                               grid_name=="Mixed agricultural fields" ~ "agriculture",
                               grid_name=="Vanilla, coffee, banana fields" ~ "agriculture",
                               grid_name=="Rice field" ~ "flooded_rice",
                               grid_name=="Vanilla" ~ "agroforest",
                               grid_name=="Village" ~ "village")) %>% 
  select(village, grid_name, season, plot, square, dead_logs, tree_dbh, tree_height, #omiting crown distance
         liana_number, herbaceous_height, herbaceous_cover, canopy_cover) #omitting observations and season

## deal with multitrunk trees by taking sqrt of sum of square DBH of each stem
habitat <- habitat %>% 
  separate(tree_dbh, c("dbh1", "dbh2", "dbh3"), sep = ",") %>% #using unique(habitat$tree_dbh) see max of 3 stems, give each their own column
  mutate(across(c(dbh1, dbh2, dbh3), as.numeric), #make numeric
         across(c(dbh1, dbh2, dbh3), ~ifelse(is.na(.x), 0, .x)), #make na into 0
         tree_dbh = dbh1^2+dbh2^2+dbh3^2, #sum sq dbhs
         tree_dbh = ifelse(tree_dbh == 0, NA, sqrt(tree_dbh))) %>% #sqrt of sum, if 0 make NA (no trees)
  select(-dbh1, -dbh2, -dbh3)

## deal with the herbaceous_height "x to x"
parse_range <- function(x) {
  x <- gsub(",", ".", as.character(x))
  unname(vapply(x, function(v) {
    if (is.na(v) || v == "NA") return(NA_real_)
    if (grepl("to", v)) {
      parts <- suppressWarnings(as.numeric(str_trim(str_split(v, "to")[[1]])))
      mean(parts, na.rm = TRUE)
    } else {
      suppressWarnings(as.numeric(v))
    }
  }, numeric(1)))
}

habitat <- habitat %>% 
  mutate(herbaceous_height = parse_range(herbaceous_height))

# add pitfall column
habitat <- habitat %>% 
  mutate(pitfall = ifelse(str_detect(plot, "Pitfall"), "Pitfall", "Grid"),
         pitfall_line = case_when(str_detect(plot, "Pitfall 1") ~ 1,
                                  str_detect(plot, "Pitfall 2") ~ 2,
                                  str_detect(plot, "Pitfall 3") ~ 3,
                                  str_detect(plot, "Pitfall 4") ~ 4,
                                  str_detect(plot, "Pitfall 5") ~ 5,
                                  str_detect(plot, "Pitfall 6") ~ 6,
                                  str_detect(plot, "Pitfall 7") ~ 7,
                                  str_detect(plot, "Pitfall 8") ~ 8,
                                  str_detect(plot, "Pitfall 9") ~ 9,
                                  str_detect(plot, "Pitfall 10") ~ 10,
                                  str_detect(plot, "Pitfall 11") ~ 11,
                                  str_detect(plot, "Pitfall 12") ~ 12,
                                  TRUE ~ 0))

## make a trees df
trees <- habitat %>% 
  select(village, grid_name, season, pitfall, plot, square, tree_height, tree_dbh) %>% 
  group_by(village, grid_name,season, plot, square) %>%
  summarise(tree_height = mean(tree_height, na.rm=T),
            tree_dbh = mean(tree_dbh, na.rm=T),
            .groups = "drop")

## create a new df individual tree measurements
squares <- habitat %>% 
  mutate(tree = ifelse(is.na(tree_dbh), 0, 1)) %>% ## count the number of trees in each square
  group_by(village, grid_name, season, plot, square) %>% 
  mutate(n_trees = sum(tree)) %>% 
  ungroup() %>% 
  select(-tree_height, -tree_dbh, -tree) %>% # remove tree info from habitat
  distinct() %>% 
  left_join(trees, by=c("village","grid_name","season","plot","square"))

# summarizing for plots
plot_sum <- squares %>% 
  mutate(across(c(n_trees, tree_height, tree_dbh, dead_logs, 
                  liana_number, herbaceous_height, 
                  herbaceous_cover, canopy_cover))) %>% 
  group_by(pitfall, pitfall_line, village, grid_name, season, plot) %>% 
  summarise(n_trees = sum(n_trees, na.rm=T),
            tree_height = mean(tree_height, na.rm=T),
            tree_dbh = mean(tree_dbh, na.rm=T),
            n_logs = sum(dead_logs, na.rm=T),
            n_liana = sum(liana_number, na.rm=T),
            m_herb_ht = mean(herbaceous_height, na.rm=T),
            m_herb_cv = mean(herbaceous_cover, na.rm = T),
            m_canopy_cv = mean(canopy_cover, na.rm = T),
            .groups = "drop") %>% 
  mutate(across(everything(), ~replace(. , is.nan(.), 0)))

# summarizing for grids
grid_sum <- plot_sum %>% 
  filter(pitfall == "Grid") %>% 
  group_by(grid_name, season, village) %>% 
  summarise_at(vars(4:11), ~mean(., na.rm=TRUE)) %>% 
  ungroup() %>% 
  arrange(grid_name) %>% 
  mutate(grid_name = factor(grid_name))

# adding village values
village_values <- grid_sum %>%
  group_by(village,season) %>%
  summarise(across(where(is.numeric), min, na.rm = TRUE), .groups = "drop") %>% 
  mutate(n_trees = 0, tree_height = 0, tree_dbh = 0, m_canopy_cv = 0, n_logs = 0, n_liana = 0) %>% 
  mutate(grid_name = "village")

# bind_rows matches columns by name (robust to column-order differences)
grid_sum <- bind_rows(grid_sum, village_values)

# name index
grid_season_name <- grid_sum %>% 
  select(grid_name, season, village) %>% 
  unite(col="grid_season", grid_name:village, remove = FALSE)

# transforming to matrix
grid_sum_mat <- grid_sum %>% 
  unite(col="grid_season", grid_name:village, remove = TRUE) %>%
  column_to_rownames("grid_season") %>%
  ungroup() %>% 
  as.matrix() 
```

```{r}
# PCA
# variables transformed to centered 0 and rescale to have unit variance before the analysis 
set.seed(123)
pca_veg <- prcomp(grid_sum_mat, center = TRUE, scale. = TRUE)

vegetation_pca <- as.data.frame(pca_veg$x) %>% 
  rownames_to_column("grid_season") %>% 
  left_join(grid_season_name, by="grid_season")

# vegetation data
x2_veg_pca <- vegetation_pca %>%
  dplyr::rename(veg_pc1 = PC1) %>%
  dplyr::rename(veg_pc2 = PC2) %>%
  dplyr::rename(grid = grid_name) %>%
  mutate(season = as.character(season)) %>% 
  dplyr::select(village, grid, season, veg_pc1, veg_pc2)


#######
# plotting

# explained variance
ex_var <- pca_veg$sdev ^2 
prop_ex_var <- ex_var/sum(ex_var)*100
sum(prop_ex_var[1:2]) 

# Convert row names to a data frame
pca_data <- data.frame(pca_veg$x[, 1:2]) %>%  # Select the first two principal components
  rownames_to_column(var = "site") %>%
  mutate(
    land_use = str_extract(site, "^[^0-9]+") %>% str_remove("_$"),  # Capture everything before the first number
    season = str_extract(site, "(?<=_)[0-9]+(?=_)"),  # Extract the number in between underscores
    village = str_extract(site, "[^_]+$")  # Capture the last part (village)
  )

pca_data$land_use <- dplyr::recode(pca_data$land_use, 
                                   "brushy_regrowth" = "Savoka",
                                   "agriculture" = "Mixed agriculture",
                                   "village" = "Village",
                                   "secondary_forest" = "Secondary forest",
                                   "flooded_rice" = "Flooded rice",
                                   "agroforest" = "Agroforest",
                                   "semi-intact_forest" = "Semi-intact forest",
                                   "village" = "Village")

pca_data$land_use <- factor(pca_data$land_use, levels = c("Semi-intact forest", "Secondary forest", "Savoka", "Mixed agriculture", "Agroforest","Flooded rice", "Village")) # Order the levels by habitat disturbance

loadings <- data.frame(pca_veg$rotation[, 1:2], variable = rownames(pca_veg$rotation)) %>%
  mutate(variable = dplyr::recode(variable,
                                  "n_trees" = "No. of trees",
                                  "tree_height" = "Avg. tree height",
                                  "tree_dbh" = "Avg. tree diameter",
                                  "n_logs" = "No. of dead logs",
                                  "n_liana" = "No. of lianas",
                                  "m_herb_ht" = "Avg. herbaceous height",
                                  "m_herb_cv" = "Avg. herbaceous cover",
                                  "m_canopy_cv" = "Avg. canopy cover"
  ))

scaling_factor <- 5

veg_colors <- c("#4d9221", "#7fbc41", "#b8e186", "#f1b6da", "#de77ae", "#c51b7d", "#8e0152")

veg_plot <- ggplot(pca_data, aes(x = PC1, y = PC2)) +
  geom_point(size = 3, aes(color = land_use, shape = village)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_segment(data = loadings, aes(x = 0, y = 0, xend = PC1*scaling_factor, yend = PC2*scaling_factor), 
               arrow = arrow(length = unit(0.2, "cm")), color = "darkgrey") +  # Vectors as arrows
  geom_text(data = loadings, aes(x = PC1*scaling_factor*1.15, y = PC2*scaling_factor*1.05, label = variable),
            color = "black", size = 3) +  # Variable names near the arrows
  labs(
    x = paste("PC1 (",round(prop_ex_var[1],2),"%)", sep = ""),
    y = paste("PC2 (",round(prop_ex_var[2],2),"%)", sep = ""),
    color = "Land-use",
    shape = "Village") +
  scale_color_manual(values = veg_colors) + 
  theme_bw() + 
  theme(aspect.ratio=1)+
  theme(axis.text=element_text(size=12,color = "black"),
        axis.title=element_text(size=14,face="bold"),
        legend.text = element_text(size = 10), 
        legend.title = element_text(size = 12, face = "bold"),
        legend.position = "right",
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1) 
  ) 


#pdf('../results/veg_plot.pdf')
veg_plot
#dev.off()

```

## Distance to the village center
```{r}
x3_ind_distance <- data_sm %>% 
  filter(host_ID %in% host_ids) %>% 
  select(village, grid, host_ID, dist_village_center) %>% 
  mutate(dist_village = as.numeric(gsub(".*?([0-9]+\\.[0-9]+|[0-9]+).*", "\\1", dist_village_center))) %>%
  mutate(dist_village = ifelse(grid == "village", 0, dist_village)) %>% 
  select(-dist_village_center, -grid)
```

## Small mammals

### PCA
```{r}
# PCA of small mammals abundance per village*grid*season

# including all small mammals in the dataset
small_mammals <- read_csv("../data/data_raw/small_mammals/Terrestrial_Mammals.csv") %>% 
  filter(grepl("TMR", animal_id)) %>% 
  filter(village == "Mandena") %>%
  dplyr::rename(host_species = field_identification, grid = habitat_type) 

# small mammals abundance per village-grid-season
dat_mat <- small_mammals %>% 
  select(host_species,village, grid, season) %>% 
  count(village, grid, season, host_species) %>% 
  spread(host_species, n, fill = 0)

# name index
grid_season_name <- dat_mat %>% 
  select(village, grid, season) %>% 
  unite(col="grid_name", village:season, remove = FALSE)

# final matrix
grid_dat_mat <- dat_mat %>% 
  unite(col="grid_name", village:season, remove = TRUE) %>% 
  column_to_rownames("grid_name") %>% 
  as.matrix()

# PCA
# I square rooted the species abundances before the analysis 
grid_dat_mat <- sqrt(grid_dat_mat)

set.seed(123)
pca_sm <- prcomp(grid_dat_mat, center = F, scale. = F)

pca_data <- as.data.frame(pca_sm$x) %>% 
  rownames_to_column(var = "site") %>%
  mutate(
    village = str_extract(site, "^[^_]+"),  # Extract everything before the first underscore (village)
    season = str_extract(site, "[0-9]+$"),  # Extract the number at the end (season)
    land_use = str_remove(site, paste0(village, "_")) %>%   # Remove village name and underscore
      str_remove(paste0("_", season, "$"))  # Remove underscore + season at the end
  ) %>%
  select(site, PC1, PC2, land_use, village, season)

# PCA scores of small-mammal community composition (used only for the descriptor
# correlations below, not as model predictors)
x4_sm_pca <- pca_data %>% 
  select(village, land_use, season, PC1, PC2) %>% 
  dplyr::rename(grid=land_use, sm_pc1=PC1, sm_pc2=PC2)

```

```{r}
# plotting

# changing names for better presentation
pca_data$land_use <- dplyr::recode(pca_data$land_use, 
                                   "brushy_regrowth" = "Savoka",
                                   "agriculture" = "Mixed agriculture",
                                   "village" = "Village",
                                   "secondary_forest" = "Secondary forest",
                                   "flooded_rice" = "Flooded rice",
                                   "agroforest" = "Agroforest",
                                   "semi-intact_forest" = "Semi-intact forest",
                                   "village" = "Village")

pca_data$land_use <- factor(pca_data$land_use, levels = c("Semi-intact forest", "Secondary forest", "Savoka", "Mixed agriculture", "Agroforest","Flooded rice", "Village"))

# explained variance
ex_var <- pca_sm$sdev ^2 
prop_ex_var <- ex_var/sum(ex_var)*100
sum(prop_ex_var[1:2]) 

# loadings
loadings <- data.frame(pca_sm$rotation[, 1:2], variable = rownames(pca_sm$rotation))

# non-native host species
non_native_sp <- c("Mus musculus", "Suncus etruscus", "Suncus murinus", "Rattus rattus")

# taking only the 5 species with the highest PC1 and PC2 loadings
loading_top <- loadings %>% 
  mutate(PC1_abs = abs(PC1)) %>%
  mutate(PC2_abs = abs(PC2)) %>%
  slice_max(n = 5, order_by = PC1_abs+PC2_abs) %>%
  mutate(status = ifelse(variable %in% non_native_sp, "Non-native", "Native"))

scaling_factor <- 5

veg_colors <- c("#4d9221", "#7fbc41", "#b8e186", "#f1b6da", "#de77ae", "#c51b7d","#a00b7d")

sm_plot <- ggplot(pca_data, aes(x = PC1, y = PC2)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_point(aes(color = land_use, shape = village), size = 3, alpha = 0.8) +
  geom_segment(data = loading_top, 
               aes(x = 0, y = 0, 
                   xend = PC1 * scaling_factor, 
                   yend = PC2 * scaling_factor, 
                   color = status),
               arrow = arrow(length = unit(0.2, "cm")),  alpha = 0.6) + 
  geom_text(data = loading_top,
            aes(PC1* scaling_factor, PC2* scaling_factor, label = variable),
            color = "black", vjust = -0.5, size = 3) +
  labs(
    x = paste("PC1 (", round(prop_ex_var[1], 2), "%)", sep = ""),
    y = paste("PC2 (", round(prop_ex_var[2], 2), "%)", sep = ""),
    color = "Land-use/Status",
    shape = "Village") +  
  scale_color_manual(values = c(veg_colors,  "darkblue", "darkred")) +
  theme_bw() + 
  theme(aspect.ratio = 1,
        axis.text = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 10), 
        legend.title = element_text(size = 12),
        legend.position = "right",
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        panel.grid = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)) 

#pdf('../results/sm_plot.pdf')
sm_plot
#dev.off()
```

## Correlations
```{r}
LU_table <- x1_host_traits %>% 
  left_join(x3_ind_distance %>% select(host_ID,dist_village), by="host_ID") %>% 
  left_join(x2_veg_pca, by=c("village","grid","season")) %>%
  left_join(x4_sm_pca, by=c("village","grid","season")) 

library(Hmisc)   # for rcorr()

cor_vars <- c("dist_village", "veg_pc1", "veg_pc2", "sm_pc1", "sm_pc2")

cor_input <- LU_table %>%
  select(all_of(cor_vars)) %>%
  as.matrix()

cor_res <- rcorr(cor_input, type = "spearman")   # Spearman correlations

cor_mat <- cor_res$r    # correlation coefficients
p_mat   <- cor_res$P    # p-values (NA on the diagonal)
n_mat   <- cor_res$n    # pairwise sample size

# the specific values that fill the [XX] placeholders (distance vs each descriptor)
others <- setdiff(cor_vars, "dist_village")
dist_cor <- tibble(
  variable = others,
  rho = cor_mat["dist_village", others],
  p   = p_mat[  "dist_village", others],
  n   = n_mat[  "dist_village", others]
)
print(dist_cor)

```


# Microbiome
Four microbiome measures are prepared. All are computed per host and z-scored *within each
species*:

* **shannon_scaled**    – Shannon alpha diversity.
* **pd_scaled**         – Faith's phylogenetic diversity (presence/absence; depth-sensitive).
* **uniqueness_scaled** – weighted-UniFrac uniqueness = mean weighted-UniFrac distance to
                          conspecifics (whole-community compositional deviation / dysbiosis).
* **FB_z**              – log10 Firmicutes/Bacteroidota ratio (a compositional log-ratio balance).

```{r}
# take the host species label from data_sm via an explicit join on host_ID
data_microbiome_filtered <- data_microbiome %>% 
  select(-host_species) %>% 
  left_join(data_sm %>% select(host_ID, host_species), by = "host_ID") %>% 
  filter(host_ID %in% host_ids)

asv_tree <- readRDS("../data/data_processed/ASV_tree_ML_rooted.rds")

metadata <- data_microbiome_filtered %>% distinct(host_ID,host_species)

# transforming into a matrix (host x ASV, relative abundances)
data_microbiome_mat <- data_microbiome_filtered %>% 
  select(host_ID, asv_ID, reads) %>% 
  pivot_wider(names_from = asv_ID, values_from = reads, values_fill = 0) %>% 
  column_to_rownames("host_ID") %>% 
  as.matrix()
```

## Shannon
```{r}
# calculating Shannon diversity
asv_shannon <- vegan::diversity(data_microbiome_mat, index = "shannon")

# table
microbiome_shannon <- data.frame(host_ID = as.numeric(rownames(data_microbiome_mat)),
                                 microbiome_shannon = asv_shannon) %>% 
  left_join(data_microbiome_filtered %>% distinct(host_ID,host_species), by="host_ID")

# standardized (z-score) by host species
microbiome_shannon_scaled <- microbiome_shannon %>%
  group_by(host_species) %>%
  mutate(
    shannon_scaled = scale(microbiome_shannon)[, 1]
  ) %>%
  ungroup() %>% 
  select(host_ID, shannon_scaled)
```

## PD Diversity
```{r}
# Pre-processing: Ensure IDs match
combined <- picante::match.phylo.comm(asv_tree,data_microbiome_mat)

# Extract the cleaned data
# This ensures any ASVs not in the tree (or vice versa) are removed
data_cleaned <- combined$comm
tree_cleaned <- combined$phy

# Calculate Phylogenetic Diversity
# include.root = TRUE uses the distance to the root of the tree
pd_results <- picante::pd(data_cleaned, tree_cleaned, include.root = TRUE)

# standardized (z-score) by host species
microbiome_pd_scaled <- pd_results %>%
  rownames_to_column("host_ID") %>% 
  mutate(host_ID = as.numeric(host_ID)) %>% 
  left_join(metadata, by="host_ID") %>% 
  group_by(host_species) %>%
  mutate(
    pd_scaled = scale(PD)[, 1]
  ) %>%
  ungroup() %>% 
  select(host_ID, pd_scaled)
```

## Uniqueness (weighted UniFrac)
```{r}
# Microbiome uniqueness = each host's mean WEIGHTED-UniFrac distance to its conspecifics.
# Computed SEPARATELY within each host species, so Rattus and Microgale are never placed
# in a shared ordination space (this is why uniqueness stays interpretable across two very
# different focal species, unlike a joint PCoA). Then z-scored per species.
#
# GUniFrac accepts the relative-abundance matrix directly (it normalizes internally).

compute_uniqueness <- function(host_ids_sp) {
  comm <- data_microbiome_mat[as.character(host_ids_sp), , drop = FALSE]
  comm <- comm[, colSums(comm) > 0, drop = FALSE]        # keep ASVs present in this species
  m  <- picante::match.phylo.comm(asv_tree, comm)        # align table & tree
  uf <- GUniFrac(m$comm, m$phy, alpha = 1)$unifracs      # alpha = 1 -> weighted UniFrac
  d  <- uf[, , "d_1"]
  diag(d) <- NA
  rowMeans(d, na.rm = TRUE)                              # names = host_ID (character)
}

# apply per species
uniq_by_sp <- lapply(
  split(metadata$host_ID, metadata$host_species),
  compute_uniqueness
)

microbiome_uniqueness <- purrr::map_dfr(
  uniq_by_sp,
  ~ tibble(host_ID = as.numeric(names(.x)), uniqueness = as.numeric(.x))
) %>%
  left_join(metadata, by = "host_ID") %>%
  group_by(host_species) %>%
  mutate(uniqueness_scaled = scale(uniqueness)[, 1]) %>%
  ungroup() %>%
  select(host_ID, uniqueness_scaled)
```

## Taxonomy
```{r}
tax <- read_delim("../data/data_raw/microbiome/ASVs_taxonomy_new.tsv") %>% 
  dplyr::rename(asv_ID = ASV)

data_microbiome_tax <- data_microbiome_filtered %>% 
  left_join(tax, by="asv_ID")

################
# Firmicutes/Bacteroidota ratio
# Summarize and spread to wide format
data_FB_ratio <- data_microbiome_tax %>% 
  group_by(host_ID, host_species, Phylum) %>% 
  summarise(reads = sum(reads), .groups = 'drop') %>% 
  filter(Phylum %in% c("Firmicutes", "Bacteroidota")) %>%
  pivot_wider(names_from = Phylum, values_from = reads, values_fill = 0)

# reads are relative abundances (0-1), so the pseudocount is on that scale.
# Use the detection limit of the filtered data (rra 0.001), and model the ratio on the log scale.
eps <- 0.001  # pseudocount = detection limit (the 0.1% relative-abundance filtering threshold)

# log-ratio, then z-score per host species
microbiome_taxa_final <- data_FB_ratio %>%
  mutate(
    FB_ratio     = (Firmicutes + eps) / (Bacteroidota + eps),
    log_FB_ratio = log10(FB_ratio)
  ) %>%
  group_by(host_species) %>%
  mutate(FB_z = as.numeric(scale(log_FB_ratio))) %>%
  ungroup() %>% 
  select(host_ID, FB_z)
```


## Microbiome final
```{r}
microbiome_final <- microbiome_shannon_scaled %>% 
  left_join(microbiome_pd_scaled, by="host_ID") %>% 
  left_join(microbiome_uniqueness, by="host_ID") %>% 
  left_join(microbiome_taxa_final, by="host_ID")
```


# Final table
Combining all variables

```{r}
x_final_table <- x1_host_traits %>% 
  left_join(x3_ind_distance %>% select(host_ID,dist_village), by="host_ID") %>% 
  left_join(microbiome_final, by="host_ID")

# check that every analysis host has complete predictors
pred_cols <- c("mass_z", "size_z", "sex", "BCI", "dist_village",
               "shannon_scaled", "pd_scaled", "uniqueness_scaled", "FB_z")
missing_pred <- x_final_table %>% filter(if_any(all_of(pred_cols), is.na))

if (nrow(missing_pred) > 0) {
  warning(nrow(missing_pred), " host(s) have missing predictors and would be lost downstream.")
  print(missing_pred %>% select(host_ID, host_species, village, grid, season, all_of(pred_cols)))
} else {
  message("All ", nrow(x_final_table), " hosts have complete predictors.")
}

# saving the table
write_csv(x_final_table, "../data/data_processed/data_x_final.csv")
```


# Random variables
Site + season + host species + individual

```{r}
rand_final_table <- x_final_table %>% 
  select(host_ID, host_species, village, grid, season) %>% 
  unite(col="site", village:grid, remove = FALSE)

# saving the table
write_csv(rand_final_table, "../data/data_processed/data_rand_final.csv")
```
